---
title: "Dummy Exam"
date: "STA 207 Part (100 points)"
output: html_document
---

<span style="color:red">**Read the instructions on Canvas carefully!**</span>

___

**Name**: Greg DePaul

**Student ID**: 917835494

___


Make sure you can load the the following package. 

```{r,echo=T,results=F,message=F}
library(lme4)
```


**(a) Dummy question: What is the date today? Use `R` to find it out and write down your answer below.**

<b> Solution: </b>  (Type your answer here)


```{r}
# (a). (Type your code in the space below, if any)
print(Sys.Date())
```
	
	
**(b) Dummy question: Write down the equation of a one-way ANOVA model with fixed effects. You can find basic LaTeX commands at the end of this file. **

<b> Solution: </b>  (Type your answer here)
	
The one-way ANOVA model will be of the form 
	
$$Y_{i,j} = \mu + \alpha_{i} + \epsilon_{i,j}$$
for 
$1 \leq i \leq r, 1 \leq j \leq n_i$ and $\epsilon_{i,j} \sim N(0, \sigma^2).$ Further, we make the assumption that 
	
$$\sum_{i = 1}^r n_i \alpha = 0$$
	
***

**LaTeX commands**

In-line: $\alpha$, $\beta$, etc.

Equation: 
$$Y_{i,j,k}, \mu, \alpha, \beta, \epsilon, i, j, H_0, H_a, \neq, =, \ldots$$

